Demography and linked selection in wild and domesticated maize

### 2015-08-12 manuscript.tex
starting in on methods

\title{Domestication and the impacts of linked selection on the maize genome}
\author{Timothy M. Beissinger\affil{1}{Dept. of Plant Sciences, University of
    California, Davis, CA, USA}\affil{2}{US Department of Agriculture, Agricultural Research Service, Columbia, MO, USA}\affil{3}{Division of Plant Sciences, University of Missouri, Columbia, MO, USA} Li Wang\affil{4}{Iowa State University, Ames, IA, USA}, Kate Crosby\affil{1}{}, Arun
  Durvasula\affil{1}{}, Matthew Hufford\affil{4}{}, \and Jeffrey
  Ross-Ibarra\affil{1}{}\affil{5}{Genome Center and Center for population biology, University of
    California, Davis, CA, USA} }

\significancetext{Patterns of linked selection, or the impact of selection on sites that neighbor a functional variant, have been carefully evaluated in only a few species. \jri{not really. we need to cite the Corbett-Detig plos-bio paper here -- he looked at a lot of species including maize!} In this work, we demonstrate that selection against deleterious mutations leaves a pronounced signature on the maize genome, reducing diversity in and immediately around genes. We show how demography interacts with selection to impact genome-wide patterns of diversity, including the important observation that rapid population expansion can increase the efficiency of selection as much as a sudden population bottleneck can weaken it. Along the way, we develop the first estimate the demographic parameters of the maize domestication from whole genome sequence data. }

@@

  \subsection{BASH, R, and Python scripts}
All analysis scripts are available in an online repository at \textcolor{red}{REPO ADDRESS HERE.}

\subsection{Plant materials}
Accessions studied were selected from the Maize HapMap2
panel \cite{chia2012, lemmon2014}. \jri{this needs slight clarification. cite chia for hapmap, but that we used higher-coverage of the teosinte from lemmon. if possible, include link to fastq or bamfiles used} Principal component analysis was employed to ensure that closely related individuals were not included due to their potential to bias results (Figure S5). Ultimately, 23 maize inbreds derived from a diverse assortment of landraces were selected for inclusion. 
Thirteen teosinte inbred lines, all members of  subspecies Z. \emph{mays}
ssp. \emph{parviglumis}, were utilized. Included lines are listed in (Table S1).
Sequences were mapped to the maize B73 version 3 reference genome \cite{schnable2009} (ftp://ftp.ensemblgenomes.org/pub/plants/release-22/fasta/zea\_mays/dna/) as described by \cite{chia2012}.\jri{did we use their bams? I thought we used bwa bams, which are not listed and weren't used for hapmap2. if that's the case, we need to mention this difference.}

\subsection{Interpolating genetic position}

Physical positions were converted to genetic positions by interpolating from the NAM genetic map \cite{glaubitz2014}, \jri{does jeff's paper really have the NAM map? I didn't think so and quick look didn't find it. i thought this came from wallace?} which provides a 1 cM resolution for physical to genetic conversion.
Within R \cite{R2014}, physical positions with corresponding genetic positions in the NAM map were used as anchors.
Physical positions in our dataset without corresponding genetic position were assigned genetic positions by scaling the anchored genetic positions according to the physical distance between the unlabeled position and the flanking anchors.

\subsection{Estimating the site frequency spectrum}
To estimate individual  and joint site frequency spectra (SFS) for maize and teosinte from inbred lines, each inbred individual was treated as representing a single haplotype from its population.
SFS were
separately computed for genic and intergenic regions, as well as for
the whole genome together. First, genic and intergenic regions were isolated using
the biomaRt package \cite{durinck2009,durinck2005} of R \cite{R2014}. Genic
regions were defined as DNA between the start and stop position of a \jri{you meant transcription start/stop not translation, right?}
gene, while intergenic regions were required to be at least 5kb up- or
down-stream from a gene start/stop. With regions defined, SFS were
estimated with ANGSD \cite{korneliussen2014}. Individual population SFS were
estimated using all positions observed in at least 80\% of the
individuals in the population, and joint SFS were estimated using all
positions observed in at least 80\% of individuals in both
populations. Individuals were
assumed to be fully inbred (-doSaf 2). Quality filters were
employed such that reads with quality score below 30 and bases with
quality score below 20 were discarded (-minMapq 30 and -minQ 20), as were reads that didn't map
uniquely (-uniqueOnly 0). Quality scores around indels were adjusted
as in Samtools (-baq 0). Genotype likelihoods were estimated using the samtools
method (-GL 1). Major and minor alleles were inferred from the data
(-doMaf 1). Because ANGSD cannot calculate a folded joint SFS, the
maize reference genome was used for polarization and then unfolded spectra
were folded using $\delta\alpha\delta{i}$ \cite{gutenkunst2009}.

\subsection{Demographic inference}
We used $\delta\alpha\delta{i}$ \cite{gutenkunst2009} to estimate the parameters of maize
domestication based on diffusion approximation of the 2-dimensional
SFS for maize and teosinte. Our first step was to specify a model, as
shown in (Figure \ref{fig:bottleneck}). In words, our model began with an ancestral teosinte
population with effective population size $N_a$ individuals. The
ancestral population split into
two distinct populations, one of maize and the other of teosinte,
$T_b$ generations in the past. The maize population immediately shrank
to size $N_b$, and then grew exponentially to size $N_{maize}$
today. From the split until today, the teosinte effective population
size remained constant. Each year, $M_{mt}$ individuals migrated from
maize to teosinte, and $M_{tm}$ individuals migrated from teosinte to
maize.The free parameters of our model were $N_a, N_b, N_{maize}, T_b,
M_{mt}, \text{and } M_{tm} $. We implimented 1,000 $\delta\alpha\delta{i}$
optimizations of this model using the ``$\delta\alpha\delta{i}$.Inference.optimize\_log\_fmin''
optimizer. In $\delta\alpha\delta{i}$ optimizations,
initial values for $N_b, N_{maize}, T_b,
M_{mt}, \text{and } M_{tm}$, respectively, were randomly perturbed up
to a factor of two from 0.02, 3, 0.04, 1e-5, and 1e-5. Lower
bounds were set at 1e-7, 1e-7,  0, 1e-10, and
1e-10, respecively, while upper bounds were respectively set at 2,
200, 1, .001, and .001. $N_a$ did not have an initial value, upper, or
lower bounds because it is estimated from equilibrium ancestral theta
($\theta = 4N_a\mu$), which $\delta\alpha\delta{i}$ determines automatically regardless
of model. Population sizes were converted from $\delta\alpha\delta{i}$'s
units using a mutation rate estimate of 3e-8
\cite{clark2005}. Among the 1,000 $\delta\alpha\delta{i}$ runs, parameters
from the run with the lowest log-likelihood were taken as estimates of
the parameters of the domestication process.

Recent population expansion, as is observed for maize, most influences
the abundance of rare alleles. However, our demographic parameter
estimation using $\delta\alpha\delta{i}$ suffered from a relatively small sample size of
23 maize individuals. Therefore, our approach likely underestimated
current maize effective population size. To address this, we utilized
a complimentary dataset of 4,021 maize landrace individuals collected
from across the Americas \textcolor{red}{(SeeDs reference here)} to
generate an independent estimate of
$N_{maize}$. These individuals were genotyped using GBS
\cite{elshire2011}, a protocol that suffers from a high rate of
missing data \cite{beissinger2013}. To minimize biases imposed from
missing data, we used R to isolate only those SNPs that
were observed in at least 1,500 individuals and subsequently projected the
SFS down to a sample of 500 individuals. Then, we utilized the singleton-based estimate of
theta ($\theta = 4N_e\mu = \frac{S}{L}$) to estimate
modern maize $N_e$ \cite{fu1993}, where $S$ is the total number of
singletons and $L$ is the total length of DNA sequenced. Based on the
GBS protocol implemented \cite{elshire2011}, $L$ is equal to the total
number of SNPs in the dataset.

>>>>

\title{Demography and linked selection in wild and domesticated maize}
\author{Timothy M. Beissinger\affil{1}{Dept. of Plant Sciences, University of
    California, Davis, CA, USA}\affil{2}{US Department of Agriculture, Agricultural Research Service, Columbia, MO, USA}\affil{3}{Division of Plant Sciences, University of Missouri, Columbia, MO, USA} Li Wang\affil{4}{Iowa State University, Ames, IA, USA}, Kate Crosby\affil{1}{}, Arun
  Durvasula\affil{1}{}, Matthew Hufford\affil{4}{}, \and Jeffrey
  Ross-Ibarra\affil{1}{}\affil{5}{Genome Center and Center for population biology, University of
    California, Davis, CA, USA} }

\significancetext{Patterns of linked selection, or the impact of selection on sites that neighbor a functional variant, have been carefully evaluated in only a few species. \jri{not really. we need to rephrase here and cite the Corbett-Detig plos-bio paper somewhere -- they looked at a lot of species including maize!} In this work, we demonstrate that selection against deleterious mutations leaves a pronounced signature on the maize genome, reducing diversity in and immediately around genes. We show how demography interacts with selection to impact genome-wide patterns of diversity, including the important observation that rapid population expansion can increase the efficiency of selection as much as a sudden population bottleneck can weaken it. Along the way, we develop the first estimate the demographic parameters of the maize domestication from whole genome sequence data. }

@@

  \subsection{BASH, R, and Python scripts}
All scripts used for analysis are available in an online repository at \textcolor{red}{REPO ADDRESS HERE.}

\subsection{Plant materials}
We made use of published sequence of inbred accessions of teosinte (Z. \emph{mays} ssp. \emph{parviglumis}) and maize landraces from the Maize HapMap2 panel \cite{chia2012, lemmon2014}. 
We removed \X teosinte and \X maize individuals that appeared as outliers in an initial principal component analysis  (Figure S5), leaving 13 teosinte and 23 maize that were used for all subsequent analyses (Table S1).

\subsection{Physical and genetic maps}
Sequences were mapped to the maize B73 version 3 reference genome \cite{schnable2009} (ftp://ftp.ensemblgenomes.org/pub/plants/release-22/fasta/zea\_mays/dna/) as described by \cite{chia2012}.\jri{did we use their bams? I thought we used bwa bams, which are not listed and weren't used for hapmap2. if that's the case, we need to mention this difference. either way, we need to include a link to the bamfiles too.}
All analyses made use of uniquely mapping reads with mapping quality score $\geq 30$ and bases with base quality score $\geq 20$, adjusting quality scores around indels. 
We converted physical coordinates to genetic coordinates via linear interpolation of the 1cM resolution NAM genetic map \cite{glaubitz2014}, \jri{i thought this should be wallace not glaubitz? need to specify what you do with physical positions before and after last marker on each chrom and markers, if any, that have physical and genetic order that disagree}.

\subsection{Estimating the site frequency spectrum}
We estimated both the genome-wide site frequency spectrum (SFS) as well as a separate SFS for genic (within annotated transcript) and intergenic ($\geq 5kb$ from a transcript) regions. 
We used the biomaRt package \cite{durinck2009,durinck2005} of R \cite{R2014} to parse annotations from the \X annotation \jri{include annotation version number and link to ENSEMBL annotation used}. 
We estimated single population and joint SFS with the software ANGSD \cite{korneliussen2014}, including all positions with at least one aligned read in $\geq 80\%$ of samples in one or both populations.
We assumed individuals were fully inbred and treated each line as a single haplotype. \jri{is doSaf -2 equivalent to using the inbreeding coefficient? in general i removed the options because they don't make sense to casual reader and the interested reader can find the script}
Because ANGSD cannot calculate a folded joint SFS, we first polarized SNPs using the maize reference genome and then folded spectra using $\delta\alpha\delta{i}$ \cite{gutenkunst2009}.

\subsection{Demographic inference}
We used the software $\delta\alpha\delta{i}$ \cite{gutenkunst2009} to estimate parameters of a domestication bottleneck from the joint maize-teosinte SFS.
We modeled an a teosinte of constant effective population size $N_a$ individuals that at time $T_b$ generations in the past gives rise to a maize population of size $N_b$ which grows exponentially to size $N_{maize}$ in the present (Figure \ref{fig:bottleneck})
The model includes migration each generation of $M_{mt}$ individuals from maize to teosinte and $M_{tm}$ individuals from teosinte to maize.  
We fixed $N_a$ using $\delta\alpha\delta{i}$'s estimation of the population mutation rate $\theta=4N_a\mu$ and a mutation rate of $3 \times 10^{-8}$ \cite{clark2005}. 
We estimated all other parameters using 1,000 $\delta\alpha\delta{i}$ optimizations of the model using ``$\delta\alpha\delta{i}$.Inference.optimize\_log\_fmin'' optimizer \jri{is this last bit about the optimizer in your code? if so, i'd ditch} setting initial parameter values at $\frac{N_b}{N_a}=0.02$, $\frac{N_{maize}}{N_a}=3$, $\frac{T_b}{2N_a}=0.04$, and $\frac{M_{mt}}{N_c}=\frac{M_{tm}}{N_a}=10^{-5}$, where the maize population size $Nc$ at time $i$ generations in the past is $N_c=N_ae^{\frac{ln(\frac{N_{maize}}{N_a})(T_b-i)}{T_b}}$.
\jri{make sure these are right (e.g. is it 2Na or 4Na for Tb?). note that we have to redo this in fractions or words because above you call Nb a population size when in reality we are estimating a proportion}
We allowed initial values to be randomly purterbed by a factor of 2, and set upper and lower bounds such that $10^{-7} \eq \frac{N_b}{N_a} \leq 2$, $10^{-7} \leq \frac{N_{maize}}{N_a} \leq{200}$, $0 \leq \frac{T_b}{2N_a} \leq 1$, $10^{-10}\leq \frac{M_{tm}}{N_a}\leq 10^{-3}$, and $10^{-10}\leq \frac{M_{mt}}{N_c}\leq 10^{-3}$
We report parameter estimates from the optimization run with the highest log-likelihood.

We additionally made use of the large genotyping data set of more than 4,000 maize landraces \citex to calculate a singleton-base estimate of the modern maize effective population size.   
We filtered these data to include only SNPs observed in $\geq 1,500$ individuals, and then projected the SFS down to a sample of 500 individuals. \jri{cite or mention how projection was done}. 
SFS down to a sample of 500 individuals. Then, we utilized the singleton-based estimate of
theta ($\theta = 4N_e\mu = \frac{S}{L}$) to estimate
modern maize $N_e$ \cite{fu1993}, where $S$ is the total number of
singletons and $L$ is the total length of DNA sequenced. Based on the
GBS protocol implemented \cite{elshire2011}, $L$ is equal to the total
number of SNPs in the dataset.
